---
title: "In-vivo evidence for increased tau deposition in temporal lobe epilepsy"
author: "Raúl R. Cruces"
date: "2025-03-21"
output:
  md_document:
    variant: gfm
always_allow_html: yes
---

# *In-vivo* evidence for increased tau deposition in temporal lobe epilepsy

### Table of Contents

1.  [Directory file content](#directory-file-content)\
2.  [Tau PET 18F-mk6240 features](#tau-pet-mk6240-features)
3.  [Data processing](#data-processing)
4.  [Data analysis](#data-analysis)

### Directory file content {#directory-file-content}

| File       | Description              |
|------------|--------------------------|
| [`Fig-1_Tau-pet_18F-mk6240_SupFig.ipynb`]    | PET-preprocessing script |
| [`Fig-2_network_contextualization.ipynb`] | notebook                 |
| [`Fig-3_clinical-cognitive_correlations.ipynb`] | notebook                 |
| [`Increased_in-vivo_tau_in_TLE.py`] | A Python script with all the                  |
| [`Increased_in-vivo_tau_in_TLE.Rmd`] | An R Markdown document includes statistical analysis and data visualization.                 |
| [`utils.py`] | A Python utility script, containing all the helper functions for the analysis                 |

# Tau PET \| 18F-mk6240 features {#tau-pet-mk6240-features}

The PET images are transformed to `NIFTI` from the `ECAT (.v)` files with `v1.0.20240202 GCC11.2.0`. This transformation keeps the data in nano Curies.

### Parameters {#parameters}

**Dimensions**: 256 x 256 x 207\
**Voxel size**: 1.21875 x 1.21875 x 1.21875\
**psfmm** : scanner PSF FWHM in mm

> psf FWHM is the full-width/half-max of the the point-spread function (PSF) of the scanner as measured in image space (also known as the burring function). The blurring function depends on the scanner and reconstruction method

### Unit reference {#unit-reference}

> cc=cm3=mL 1 nCi = 37 Bq\
> Bq: Becquerels\
> nCi: nano Curies

### 3D non motion corrected data

1.  `TX256.v` Linear atenuation map, 4D_MC is corregistered to this image. Later this is the image that is use to calculate the affine registration between PET and MRi space. \> This is a CT transmission scan
2.  `EM_3D.v` 4 frames, 20 minutes of acquisition each one (Bq/cc).
3.  `EM_3D_AVG.v` average of the four frames (Bq/cc).

### 4D motion corrected data

1.  `EM_4D_MC01.v` Filter Back Projection (FBP) image. It is the inversion of the radon transformation. 4 frames, 20 minutes of acquisition each one (nCi/cc).
2.  `EM_4D_MC01_AVG.v` FBP average of 4 four frames (nCi/cc units).

### Standardized uptake values

In summary this gives the following equation to calculate SUV at time $t$ post injection:

> ![SUV](https://latex.codecogs.com/svg.image?%7B\color%7BWhite%7D&space;SUV(t)&space;=&space;\frac&space;%7Bc_%7Bimg%7D(t)%7D&space;%7BID&space;/&space;BW%7D%7D)\
> **Cimg**=PET image\
> **ID**=Injected dose\
> **BW**=body weight\
> With the radioactivity measured from an image acquired at (or around) the time t, decay corrected to t=0 and expressed as volume concentration (e.g. MBq/mL), the injected dose ID at t=0 (e.g. in MBq), and the body weight BW (near the time of image acquisition) implicitly converted into the body volume assuming an average mass density of 1 g/mL.

**SUVR (ratio)**: The injected activity, body weight, and mass density, which are all components of the SUV calculation, cancel each other out.

> ![](https://latex.codecogs.com/svg.image?%7B\color%7BWhite%7D&space;%7B\mathit%7BSUVR%7D%7D&space;=&space;\frac&space;%7B\mathit%7BSUV_%7Btarget%7D%7D%7D&space;%7B\mathit%7BSUV_%7Breference%7D%7D%7D&space;=&space;\frac&space;%7B\mathit%7Bc_%7Bimg,target%7D%7D%7D&space;%7B\mathit%7Bc_%7Bimg,reference%7D%7D%7D%7D)

# Analysis {#data-analysis}

## Database description

| Column name            | Description                                                 |
|------------------------|------------------------------------------------|
| `EpiTrack.Score.age.corrected` | Composite score of the EpiTrack battery.            |
| `Epi.Acc.D.per`        | Accuracy in the delayed recall task.                        |
| `task.semantic`        | Accuracy in the semantic task.                              |
| `lateralization`       | Laterality of the seizure onset in TLE (R, L)               |
| `mCi.net`              | Net injected dose of the radiotracer.                       |
| `mri.Tdiff`            | Time difference between the MRI and the PET scan.           |
| `group`                | Diagnostic group (Healthy, Patient).                        |
| `mk6240.session`       | Session of the PET scan (1,2).                              |
| `mk6240.sig.ipsi`      | Mean 18F-mk6240 SUVR in the ipsilateral significant areas   |
| `mk6240.sig.contra`    | Mean 18F-mk6240 SUVR in the contralateral significant areas |

### Setup the environment

```{r setup, include=FALSE, warning=FALSE}

# All chunks in the document will display their code
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
# Load libraries
library("lubridate") # interval
library("dplyr") # %>%
library("gtsummary") # Tables reports
library("kableExtra") # tables formatting 
library("scales") # alpha
library("ggplot2") # lollipop
library("lme4") # mixed-effects models
library("lmerTest") # mixed-effects models
library("pander") # table format
library("osfr")  # OSF interface
library("gridExtra") # grid.arrange
library("emmeans")  # For estimating marginal means

```

### Get the data from the OSF repository

```{r data, echo=FALSE, message=FALSE, warning=FALSE}
# create a temporary directory
tmpdir= tempdir()

# Download the csv database
osf_file <- osf_retrieve_node("ct3gw") %>%
  osf_ls_files %>%
  filter(name == "18F-MK6240_in-vivo_database.csv") %>%
  osf_download(path = tmpdir,conflicts = "overwritte")

# Load the csv database
mk.df <- read.csv(osf_file$local_path)

# Remove the tom variables
rm(osf_file, tmpdir)
```

## Participants

```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
# Mean and Std
mk.df %>%
  dplyr::select(group, handedness, age, sex, mk6240.session, mCi.net, mk6240.Tdiff, EpiTrack.Score.age.corrected, Epi.Acc.D.per, task.semantic, hip.ipsi, hip.cntr) %>%
  tbl_summary(by = group, missing = "no",
              statistic = list(all_continuous() ~ "{mean}±{sd}"),
              type = list(mCi.net ~ "continuous", age~"continuous"),
              digits = list(age~1)) %>%
  modify_header(label = "**Demographics**") %>% 
  add_p(test = all_continuous() ~ "t.test") %>%
  # adding the header will also unhide the column
  modify_header(statistic ~ "**Statistic**") %>%
    as_kable_extra(
      booktabs = TRUE,
      longtable = TRUE,
      linesep = ""
    ) %>%
    kableExtra::kable_styling(
      position = "left",
      latex_options = c("striped", "repeat_header"),
      stripe_color = "gray!15"
    )
```

## 18F-MK6240 SUVR

```{r table2, echo=FALSE, message=FALSE, warning=FALSE}
# Mean and Std
mk.df %>%
  dplyr::select(group, mk6240.mean,  mk6240.mean.ipsi, mk6240.mean.contra, mk6240.sig, mk6240.sig.ipsi,  mk6240.sig.contra) %>%
  tbl_summary(by = group, missing = "no",
              statistic = list(all_continuous() ~ "{mean}±{sd}"),
              type = list(mk6240.mean ~ "continuous", mk6240.mean.ipsi~"continuous")) %>%
  modify_header(label = "**18F-MK6240 SUVR**") %>% 
  # add a header to the statistic column, which is hidden by default
  add_p(test = all_continuous() ~ "t.test") %>%
  # adding the header will also unhide the column
  modify_header(statistic ~ "**Statistic**") %>%
    as_kable_extra(
      booktabs = TRUE,
      longtable = TRUE,
      linesep = ""
    ) %>%
    kableExtra::kable_styling(
      position = "left",
      latex_options = c("striped", "repeat_header"),
      stripe_color = "gray!15"
    )
```

```{r table3, echo=FALSE, message=FALSE, warning=FALSE}
# Mean and Std
mk.df %>%
  subset(group=="Patient") %>% 
  dplyr::select(group, duration, lateralization, origin) %>%
  tbl_summary(by=group, missing = "no",
              statistic = list(all_continuous() ~ "{mean}±{sd}"),
              type = list(duration ~ "continuous"),
              digits = list(duration~1)) %>%
  modify_header(label = "**Patients**") %>% 
    as_kable_extra(
      booktabs = TRUE,
      longtable = TRUE,
      linesep = ""
    ) %>%
    kableExtra::kable_styling(
      position = "left",
      latex_options = c("striped", "repeat_header"),
      stripe_color = "gray!15"
    )
```

## Figure1.B \| Group Differences in MK-6240 Uptake.

Violin plots display the mean SUVR values for significant regions in each hemisphere by group, with significant differences assessed using a two-tailed Wilcoxon rank-sum test.

```{r figure1b.ipsi, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=4}
# Function to produce summary statistics (mean and +/- sd)
# > Saved as SVG 450 x 400 px
data_summary <- function(x) {
  m <- mean(x)
  ymin <- m-sd(x)
  ymax <- m+sd(x)
  return(c(y=m,ymin=ymin,ymax=ymax))
}

# Reorder levels
mk.df$group <- factor(mk.df$group, levels = c("Patient", "Healthy"))

# Ipsilateral values per type
g.ipsi <- ggplot(mk.df, aes(x=group, y=mk6240.sig.ipsi, color=group)) + 
  geom_violin() + 
  geom_jitter(shape=16, position=position_jitter(0.1)) + 
  stat_summary(fun.data=data_summary) + 
  coord_cartesian(ylim = c(0.8, 1.8)) +
  ggtitle("Ipsilateral") +
  scale_color_manual(values=c("#ff5555", "#666666")) +
  labs(y = "Mean significant SUVR") +
  theme(plot.title = element_text(hjust = 0.5),
    legend.position = "none")

# Contralateral values per type
g.contra <- ggplot(mk.df, aes(x=group, y=mk6240.sig.contra, color=group)) + 
  geom_violin() + 
  geom_jitter(shape=16, position=position_jitter(0.1)) + 
  stat_summary(fun.data=data_summary) + 
  coord_cartesian(ylim = c(0.8, 1.8)) +
  ggtitle("Contralateral") +
  scale_color_manual(values=c("#ff5555", "#666666"))  +
  labs(y = "Mean significant SUVR") +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = c(0.8, 0.8),  # Adjust legend inside the plot
    legend.background = element_rect(fill = "transparent", color = NA)  # Transparent legend box
  )

# Arrange plots in a grid
grid.arrange(grobs = list(g.ipsi, g.contra), ncol = 2)

```

## Figure.3A \| Tau MK-6240 SUVR and clinical relationships

Scatter plot display of the relationship of mean MK-6240 SUVR with behavioral and clinical measures. Duration and age are measured in *years*, while all the behavioral measurements where z-scores based in the control group.

```{r figure3a, fig.width=6.5, fig.height=3, warning=FALSE, echo=FALSE, message=FALSE}

# FUNCTION - Plot correlations using ggplot2
plot_corr_ggplot <- function(mk620, value, Col=NULL, Xlab="Z-score", Ylab="mean SUVR",
                             ln.col="gray65", ci.col="lightgray", Ylim=c(0.75,1.5), sub=NULL, Title=NULL) {
  # Calculate correlation
  cmod <- cor.test(value, mk620, method = "pearson")
  rval <- round(cmod$estimate, 2)
  rsig <- round(cmod$p.value, 2)
  
  # Create data frame for ggplot
  df <- data.frame(value = value, mk620 = mk620, Col = Col)
  
  # Create ggplot
  p <- ggplot(df, aes(x = value, y = mk620)) +
    geom_point(aes(color = Col), size = 3) +
    geom_smooth(method = "lm", se = TRUE, color = ln.col, fill = ci.col) +
    labs(title = paste(Title, "\n", "r=", rval, ", pval=", rsig),
         x = Xlab, y = Ylab) +
    ylim(Ylim) +
    scale_color_manual(values = c("#ff5555", "#666666")) +
    theme_minimal() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5))
  
  return(p)
}


# Subset of cross-sectional subjects
mk.ses1 <- mk.df[mk.df$mk6240.session == 1, ]

# Clinical correlations with mean pet deposits
tle <- which(mk.ses1$group == "Patient")

# Colors as Longitudinal subjects
ind <- as.character(mk.ses1$participant_id) %in% mk.ses1$participant_id[duplicated(mk.ses1$participant_id)]

# Color vector
col.group <- c("#666666", "#ff5555")[mk.ses1$group]

# Variables to correlate
corr.cli <- c("age", "duration")

# List to store plots
plot_list <- list()

# Set the variable to correlate with clinical
for (i in seq_along(corr.cli)) {
  var_name <- corr.cli[i]
  for (j in c("mk6240.mean")) {
    mk.val <- mk.ses1[, j]
    
    # Plot data
    x <- mk.ses1[, var_name]
    p <- plot_corr_ggplot(mk.val, x, Xlab = "years", Ylab = "Mean SUVR", Col = col.group, sub = mk.ses1$participant_id, Title = corr.cli[i], Ylim = c(0.8, 1.5))
    plot_list <- c(plot_list, list(p))
  }
}

# Arrange plots in a grid
grid.arrange(grobs = plot_list, ncol = 2)

```

## Figure.3B \| Tau MK-6240 SUVR and behavioural relationships

# Correlation with behavior 1000 x 945 px

```{r figure3b, fig.width=10, fig.height=3, warning=FALSE, echo=FALSE, message=FALSE}

# Variables to correlate
corr.cli <- c("EpiTrack.Score.age.corrected", "Epi.Acc.D.per", "task.semantic")

# List to store plots
plot_list <- list()

# Set the variable to correlate with clinical
for (i in seq_along(corr.cli)) {
  var_name <- corr.cli[i]
  for (j in c("mk6240.sig")) {
    mk.val <- mk.ses1[, j]
    
    # Plot data
    x <- mk.ses1[, var_name]
    p <- plot_corr_ggplot(mk.val, x, Xlab = "z-score", Ylab = "Mean significant SUVR", Col = col.group, sub = mk.ses1$participant_id, Title = corr.cli[i], Ylim = c(0.8, 1.8))
    plot_list <- c(plot_list, list(p))
  }
}

# Arrange plots in a grid
grid.arrange(grobs = plot_list, ncol = 3)

```

## Supplementary Figure.4 \| Mean longitudinal *18F-MK6240*

Differential trajectories of $[F^18]MK-6240$ uptake by group between the PET scan session 1 and 2 in the significant areas. The lines correspond to the longitudinal subjects.

```{r Supfigure4, fig.width=10, fig.height=5, warning=FALSE, echo=FALSE, message=FALSE}
plot_long <- function(cmk.df, col_name="mk6240.sig", cols=c("#ff5555", "#666666"), Title="") {
  # Colors as Longitudinal subjects
  
  # Get the index of subjects with mk6240.session 2
  ind <- as.character(cmk.df$participant_id) %in% cmk.df$participant_id[duplicated(cmk.df$participant_id)]
  
  # Create a cvector of color by type
  sub.uni <- cols[as.factor(cmk.df$group)]
  tle.uni <- ifelse(ind==TRUE, sub.uni, alpha(sub.uni,0.4))
  
  x <- jitter(cmk.df$mk6240.session, factor = 0.5)
  plot(x, cmk.df[,col_name], pch=20, bty='n', cex=3, col=tle.uni, xlab = "Session", 
      ylab = "mk6240 - mean SUVR", xaxt = "n", ylim = c(0.8,1.8), main=Title)
  # Y-axis
  axis(1, at = c(1, 2))
  
  
  df.long <- as.data.frame(cmk.df)
  df.long$x.jt <- x
  df.long$col <- cols[as.factor(cmk.df$group)]
  
  for (k in unique(df.long$participant_id)) {
      xx <- df.long[df.long$participant_id==k, "x.jt"]
      yy <- df.long[df.long$participant_id==k, col_name]
      if (length(yy)==2){
        lines(xx, yy, col=df.long$col[df.long$participant_id==k])
      }
    }
}

# Set the layout 1 row 2 columns
par(mfrow=c(1,2))


# Plot Patients
plot_long(mk.df, col_name = "mk6240.sig", cols=c("#ff5555", NA), Title="Patients")

# Plot Healthy controls
plot_long(mk.df, col_name = "mk6240.sig", cols=c(NA, "#666666"), Title="Healthy controls")

```

```{r lme, warning=FALSE, echo=FALSE, message=FALSE}

# Make group a factor
mk.df$group <- as.factor(mk.df$group)

# Change months to years
mk.df$mk6240.Tdiff.yrs <- mk.df$mk6240.Tdiff/12

# Fit the mixed-effects model
mem <- lmer(mk6240.sig ~ mk6240.Tdiff.yrs * group + (1 | participant_id), data = mk.df)

# Summarize the model to view the fixed effects
summary(mem)

# Calculate the estimated marginal means for the slopes (change per year)
emtrends_result <- emtrends(mem, pairwise ~ group, var = "mk6240.Tdiff.yrs")

# View the estimated slopes by group
print(emtrends_result$emtrends)
```

No significant interaction (`mk6240.Tdiff.yrs:groupHealthy`, p = 0.525) was found, suggesting that the rate of change in `mk6240.sig` over time is not significantly different between the Healthy and Patient groups.

-   **Healthy Group**: The estimated slope is `-0.0088` per year (95% CI: -0.0425 to 0.0249). This suggests a slight decrease in `mk6240.sig` over time, but the change is small and not statistically significant (p = 0.704).

-   **Patient Group**: The estimated slope is `0.0078` per year (95% CI: -0.0347 to 0.0502), indicating a slight increase in `mk6240.sig` over time, though this change is also not statistically significant.

There is no strong evidence that `mk6240.sig` changes over time in either the Healthy or Patient groups, nor that the rate of change differs significantly between them. However, the significant main effect of group (p < 0.001) suggests that `mk6240.sig` is, on average, lower in the Healthy group compared to Patients.


```{r, include=FALSE}
   # add this chunk to end of mycode.rmd
   file.rename(from="Increased_in-vivo_tau_in_TLE.md", 
               to="README.md")
```
